\input{content/macros.tex}

\PanelTitle{Definition 4.1 (Lyapunov stability)}

Consider the autonomous system

\[
\dot{x} = f(x), \qquad f(0)=0
\]

so that the origin is an equilibrium point.

The equilibrium is said to be \textbf{stable in the sense of Lyapunov} if

\[
\forall \varepsilon > 0 \; \exists \delta(\varepsilon) > 0
\]

such that

\[
\|x(0)\| < \delta \Rightarrow \|x(t)\| < \varepsilon
\quad \forall t \ge 0 .
\]

\textbf{1. Geometric meaning}

The definition compares two balls centered at the equilibrium.

\begin{itemize}
\item an outer ball of radius $\varepsilon$
\item an inner ball of radius $\delta$
\end{itemize}

The property says that if the trajectory starts inside the $\delta$
ball, it will never leave the $\varepsilon$ ball.

Thus stability is a \textbf{confinement property}.  
The trajectory may move inside the $\varepsilon$ region, but it cannot
escape it.

\textbf{2. The order of quantifiers}

The most important part of the definition is the order

\[
\forall \varepsilon >0 \quad \exists \delta(\varepsilon) >0 .
\]

This means the following.

First the challenger chooses $\varepsilon$.

Then we must produce a value of $\delta$ that guarantees the property
for that value of $\varepsilon$.

The dependence $\delta=\delta(\varepsilon)$ is therefore essential.

If the order were reversed,

\[
\exists \delta>0 \quad \forall \varepsilon>0
\]

the definition would be impossible to satisfy.

\textbf{3. What stability does not require}

The definition does not require trajectories to approach the
equilibrium.

A trajectory may orbit forever around the equilibrium and still be
stable.

A typical example is a frictionless pendulum, whose trajectories are
closed orbits.

\textbf{4. Asymptotic stability}

An equilibrium is \textbf{asymptotically stable} if

\begin{itemize}
\item it is Lyapunov stable
\item trajectories starting sufficiently close satisfy
\end{itemize}

\[
\lim_{t\to\infty} x(t)=0 .
\]

Thus asymptotic stability combines

\[
\text{confinement} + \text{convergence}.
\]

\textbf{5. Why the definition requires $t\ge0$}

The condition

\[
\|x(t)\| < \varepsilon \quad \forall t \ge 0
\]

means that the trajectory must remain inside the $\varepsilon$ region
for all future time.

If the condition held only for a finite time interval, the trajectory
could eventually escape.

\textbf{6. Interpretation}

Lyapunov stability answers the question:

\emph{If the system is slightly perturbed from equilibrium, will the
state remain close to it?}

This notion is fundamental because real systems are never initialized
exactly at equilibrium.
